---
title: "CCMPP Input Data Structures Quick Guide"
author: ""
date: |
    | `r Sys.Date()`
    | (`r paste0("_ccmppWPP_ version ", packageVersion("ccmppWPP"), ")")`
# output:
#     function(...) bookdown::html_document2(..., base_format = rmarkdown::html_vignette, css=c(system.file("rmarkdown", "templates", "html_vignette", "resources", "vignette.css", package = "rmarkdown"), "vignette2.css")):
#       toc: true
#       toc_depth: 2
output:
  word_document:
    reference_docx: docx_reference.docx    
bibliography: bibliography.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{CCMPP Input Data Structures Quick Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  results = TRUE
)

## set global chunk options
options(formatR.arrow=TRUE,width=80)
```

<!-- 
STYLE CONVENTIONS
- Filenames: single quoted
- Software: italics
- S3 classes: italics
- Object attributes: double quoted
-->


```{r source-package-functions, echo = FALSE, results = "hide"}
library(ccmppWPP)
```


# Introduction

This is a quick guide to using the S3 classes for CCMPP input objects. See the vignette "S3 Classes for CCMPP Input Data Structures" for full details of the classes.


# The S3 Classes for CCMPP

## Role in CCMPP Workflow

`ccmppWPP_workflow_one_country_variant` is the workhorse function for performing the CCMPP on a single country. It takes a single argument, a list of data frame, each containing data for a component of demographic change (population at baseline, fertility, etc.). The S3 class `ccmpp_input_list` was created to 

  1. Rigorously define the input list (names of elements, structure of the data frames);
  2. Encode the definition in *R* through a set of *constructor* and *validator* functions.
  
The constructor functions provide a way to create valid inputs to `ccmpp_input_list` from data frames. The validator functions provide a mechanism for validating arbitrary lists as members of the `ccmpp_input_list` class.

A single line is added to `ccmppWPP_workflow_one_country_variant` to enact the validation:

```{r, eval = FALSE}
ccmppWPP_workflow_one_country_variant <- function(wpp_input) {

    wpp_input <- as_ccmpp_input_list(wpp_input) #<<<<<<<< ADDED: VALIDATE INPUTS

    # extract objects needed for ccmpp
    ccmpp_input <- data_parse_ccmpp_input(indata = wpp_input)

    ... etc.
    ## rest of code.. 
```


## Structure of CCMPP Input Lists

Table: (\#tab:ccmpp-input-subclass-restrictions) Names, demographic dimensions, and value types of data frame components of `ccmpp_input_list` objects.

| Element name              | Dimensions                | Value type  | Value scale |
|:--------------------------|:--------------------------|:------------|:------------|
| `fert_rate_age_f`         | time, age                 | rate        | numeric     |
| `srb`                     | time                      | ratio       | `NA`        |
| `survival_ratio_input_df` | time, sex, age            | proportion  | `NA`        |
| `mig_net_rate_age_sex`    | time, sex, age            | rate        | numeric     |
| `mig_net_count_age_sex`   | time, sex, age            | count       | numeric     |
| `mig_net_count_tot_b`     | time                      | count       | numeric     |
| `mig_parameter`           | time                      | categorical | `NA`        |
| `pop_count_age_sex_base`  | time, sex, age            | count       | numeric     |
| `life_table_age_sex`      | indicator, time, sex, age | real        | numeric     |

The elements of a `ccmpp_input_list` object are in Table \@ref(tab:ccmpp-input-subclass-restrictions). Each is a data frame with column names taken from the following list:

  - indicator
  - time_start
  - time_span
  - age_start
  - age_span 
  - sex
  - value
  
Each data frame is also given an S3 class with the same name as its element. The element `fert_rate_age_f` is a data frame with class `fert_rate_age_f`, and so on. These data frames have *attributes* (meta-data), including "dimension", "value_type", and "value_scale". The "dimension" attribute indicates the demographic breakdown of the data. Use the function `demog_change_component_dimensions()` on an object to extract them. Allowed dimensions are 

  - indicator
  - time
  - sex
  - age
  
The "value_type" attribute indicates the type of values the `value` column contains. Use the function `value_type()` on an object to extract it. Allowed "value_type"s are 

  - count
  - rate
  - ratio
  - proportion
  - percentage
  - real
  - categorical

Type "real" is a catch-all for numeric values that do not fall under any other type. It is also used when the `value` column contains a mixture of the other types, as is the case for `life_table_age_sex` objects. 

The "value_scale" attribute specifies the scale of the `value` column. Use the function `value_scale()` on an object to extract it. It is either a numeric scalar (`1` by default) or `NA`. It is non-missing only for objects with "value_type"s `r toString(ccmppWPP:::get_value_types_w_non_NA_value_scale())`. Objects of class `life_table_age_sex` are the exception where the default is `1e+05` to indicate the radix. The interpretation of the "value_scale" is dependent on the both the "value_type" and object class. For "value_types" `count` and `real` "value_scale" is a mulitplier. For `count` a typical "value_type" might be `1000` to indicate the counts are in thousands. For "value_type" "rate" the "value_scale" is interpreted as the scale of the denominator (exposure).


## Creating Data Frame Components

To create a data frame for inclusion in a `ccmpp_input_list`, use the constructor function of the same name:

```{r}
library(ccmppWPP) #load the package first
data(wpp_input_example) #load the example data

## The 'fert_rate_age_f' data frame in the example is a basic R data frame:
class(wpp_input_example$fert_rate_age_f)

## Use the constructor to create a data frame with extra class memberships:
fert_example <- fert_rate_age_f(wpp_input_example$fert_rate_age_f)
class(fert_example)

## Look at the attributes of the new object
names(attributes(fert_example))
```

The classed data frame has the standard data frame attributes, "names", "row.names", "class", and "dimensions". Additionally, it has attributes "dimensions", "value_type", 
"age_span", "time_span", and "non_zero_fert_ages". These can be inspected with functions of the same name, e.g.,:

```{r}
age_span(fert_example)
non_zero_fert_ages(fert_example)
```

Special `print` and `summary` methods have been written to concisely display the data frames:

```{r}
## Print method
fert_example
```

```{r}
## Summary method
summary(fert_example)
```

Similar functions exist for the other data frame elements. 


## Validating Data Frame Components

At any time, the validity of an object as a member of its class can be checked using the function `validate_ccmpp_object`. This function inspects the objects "class" and calls the appropriate subfunction to check that the data structure and attributes of the object match the class definition. If they match the object is simply returned as-is; this is standard *R* convention. If the object is not valid, an error is signalled:

```{r}
## Not run:
## validate_ccmpp_object(wpp_input_example$fert_rate_age_f)
## Error in validate_ccmpp_object.default(wpp_input_example$fert_rate_age_f) : 
##   'x' is not an object with a valid CCMPP object class. 'class(x) = data.frame'. Valid classes are 'life_table_age_sex', 'mig_parameter', 'mig_net_count_tot_b', 'mig_net_count_age_sex', 'mig_net_rate_age_sex', 'srb', 'pop_count_age_sex_base', 'survival_ratio_input_df', 'fert_rate_age_f', 'ccmpp_input_df', 'demog_change_component_df'.

## Valid objects are simply returned to the console
validate_ccmpp_object(fert_example)
```


## Creating Input Lists

### From Data Frames

The function `ccmpp_input_list` can be used to create a valid input list from constituent data frames:

```{r}
ccmpp_input_list_example <-
    ccmpp_input_list(pop_count_age_sex_base = wpp_input_example$pop_count_age_sex_base,
                     life_table_age_sex = wpp_input_example$life_table_age_sex,
                     fert_rate_age_f = wpp_input_example$fert_rate_age_f,
                     srb = wpp_input_example$srb,
                     mig_net_count_age_sex = wpp_input_example$mig_net_count_age_sex,
                     mig_net_rate_age_sex = wpp_input_example$mig_net_rate_age_sex,
                     mig_net_count_tot_b = wpp_input_example$mig_net_count_tot_b,
                     mig_parameter = wpp_input_example$mig_parameter)

class(ccmpp_input_list_example)
```


### From a List

If the user has already assembled a list with the necessary components and proper element names, `as_ccmpp_input_list` can be used to *coerce* it to a member of the class:

```{r}
ccmpp_input_list_example_2 <-
    as_ccmpp_input_list(wpp_input_example)

class(ccmpp_input_list_example_2)
```

       
## Validating Input Lists

The same function used above, `validate_ccmpp_object` can be used to validate a list

```{r}
validate_ccmpp_object(ccmpp_input_list_example)
```


# Working with CCMPP Input Classes

In general, `ccmpp_input_list`s and the component data frames will behave just like regular *R* lists and data frames. However, it is useful to note that subsetting a data frame component of a `ccmpp_input_list` will drop the special class and the attributes. This is because a data frame must have exactly the right columns, column names, and structure to be a valid member of the class. Subsetting or otherwise manipulating a data frame will likely make it invalid as a member.

```{r}
class(fert_example)
```
```{r}
x <- fert_example[1:5,]
class(x)
attributes(x)
```

```{r}
y <- fert_example$time_start
class(y)
attributes(x)
```

To subset a data frame there are special functions `subset_[dimension]`, where "`[dimension]`" is one of the dimensions in Table \@ref(tab:ccmpp-input-subclass-restrictions). 

```{r}
fert_example_1950s_only <- subset_time(fert_example, times = 1950:1959)
summary(fert_example_1950s_only)
```

Alternatively, subset the object as normal and recreate it as a member of the class:

```{r}
### Tidyverse style:
library(magrittr)
library(dplyr)
fert_example_1950s_only_2 <-
    dplyr::filter(fert_example, time_start < 1960) %>%
    fert_rate_age_f()
class(fert_example_1950s_only_2)

### Base R style:
fert_example_1950s_only_3 <-
    fert_example[fert_example$time_start < 1960, ]
fert_example_1950s_only_3 <- fert_rate_age_f(fert_example_1950s_only_3)
class(fert_example_1950s_only_3)
```


# What Next?

1. Get feedback on the general set-up, names of things, operation, etc.
2. Decide if the output list would benefit from having classes defined like the input list. This would be straightforward given the infrastructure already created. 
3. Complete documentation files for functions and classes (e.g., `?ccmpp_input_list`).
4. Consider other methods that would be useful. E.g., would a `plot` method be useful to vizualize the inputs? This would allow quick and easier creation of a standard set of vizualizations to check the inputs/outputs.

